---
title: "Portugal Tourism Forecast"
author: ' Group Project'
date: "30 Oct 2025"
output:
  html_document:
    df_print: paged
link-citations: true
nocite: '@*'
references:
- id: "oecd-guidance-2025"
  type: report
  title: Guidance to develop a tourism demand forecasting model for Portugal
  author:
  - literal: OECD
  issued:
    date-parts:
    - - 2025
      - 5
      - 30
  container-title: OECD Tourism Papers
  collection-number: 2025/10
  publisher: OECD Publishing
  publisher-place: Paris
  DOI: "10.1787/8a941e5e-en"
  URL: "http://dx.doi.org/10.1787/8a941e5e-en"
- id: "eurostat-pt-nights"
  type: dataset
  title: "Arrivals at tourist accommodation establishments - monthly data — Portugal"
  author:
  - literal: Eurostat
  issued:
    date-parts: 2025
  publisher: Statistical Office of the European Union
  publisher-place: Luxembourg
  URL: https://ec.europa.eu/eurostat
- id: "fpp3-book"
  type: book
  title: 'Forecasting: Principles and Practice (3rd ed.)'
  author:
  - family: Hyndman
    given: Rob J.
  - family: Athanasopoulos
    given: George
  issued:
    date-parts: 2021
  publisher: OTexts
  URL: https://otexts.com/fpp3/
- id: "r-core"
  type: software
  title: 'R: A Language and Environment for Statistical Computing'
  author:
  - literal: R Core Team
  issued:
    date-parts: 2025
  publisher: R Foundation for Statistical Computing
  publisher-place: Vienna, Austria
  URL: "https://www.r-project.org/"
- id: "pkg-tsibble"
  type: software
  title: 'tsibble: Tidy Temporal Data Frames and Tools'
  author:
  - family: Wang
    given: Earo
  - family: Cook
    given: Dianne
  - family: Hyndman
    given: Rob J.
  issued:
    date-parts: 2024
  URL: https://tsibble.tidyverts.org/
- id: "pkg-feasts"
  type: software
  title: 'feasts: Feature Extraction and Statistics for Time Series'
  author:
  - family: Hyndman
    given: Rob J.
  - family: "O'Hara-Wild"
    given: Mitchell
  - family: Wang
    given: Earo
  issued:
    date-parts: 2024
  URL: https://feasts.tidyverts.org/
- id: "pkg-urca"
  type: software
  title: 'urca: Unit Root and Cointegration Tests for Time Series Data'
  author:
  - family: Pfaff
    given: Bernhard
  issued:
    date-parts: 2024
  URL: "https://cran.r-project.org/package=urca"
- id: "pkg-readr"
  type: software
  title: 'readr: Read Rectangular Text Data'
  author:
  - family: Wickham
    given: Hadley
  - family: Hester
    given: Jim
  - family: Bryan
    given: Jennifer
  issued:
    date-parts: 2024
  URL: https://readr.tidyverse.org/
- id: "pkg-kableextra"
  type: software
  title: 'kableExtra: Construct Complex Table with ''kable'' and Pipe Syntax'
  author:
  - family: Zhu
    given: Hao
  issued:
    date-parts: 2024
  URL: https://haozhu233.github.io/kableExtra/
- id: "pkg-prettydoc"
  type: software
  title: 'prettydoc: Creating Pretty Documents from R Markdown'
  author:
  - family: Tang
    given: Yixuan
  issued:
    date-parts: 2024
  URL: "https://cran.r-project.org/package=prettydoc"
editor_options:
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##  Introduction

Tourism in Portugal is a major economic driver. This makes it an excellent data-set for a forecasting case study. According to OECD Tourism Papers 2025/10, travel and tourism made up nearly half of Portugal’s service exports and directly contributed about 9.7% of national GDP in 2024. This industry also recorded 80.3 million overnight stays in 2024, which is a 4.8% increase from the previous year. 
Stakeholders at all levels need projections of visitor volumes to make well-informed choices. The monthly Portuguese tourism data-set from Euro Stat (www.eurostat.com) is particularly ideal in this case. It is an official, high-quality time series that provides timely insights into travel demand trends and offering extensive historical data. Monthly data-set is a good choice because it captures clear fluctuations and seasonal patterns that might otherwise be missed in less frequent data. Portugal’s tourism demand shows a clear seasonal trend. These regular highs and lows make the series ideal for Seasonal ARIMA (SARIMA) modeling. The SARIMA framework is well-suited to capture periodic seasonality along with overall trends. 
Forecasting monthly tourism volumes in Portugal can provide significant policy, economic, and operational benefits. Government agencies and tourism authorities can predict demand spikes or declines, allowing for better resource allocation and infrastructure planning, such as preparing transportation networks and public services for busy tourist seasons. Businesses in the travel and hospitality sector rely on these forecasts for staffing and inventory management. This ensures they have enough employees, hotel rooms, flights, and other services available to meet expected demand. Summarily, accurate tourism demand forecasts help stakeholders optimize operations and budgets, prevent overcrowding during peak times, and create sustainable tourism policies.

---

### Installation of all required packages
```{r message = FALSE, warning = FALSE}
library("fpp3")
library("kableExtra")
library("tsibble")
```

### Loading the data and dealing with the missing values
```{r message = FALSE, warning = FALSE}
library(readr)
pt_tourism_dataset <- read_csv("pt_tourism.csv")
pt_tourism_dataset <- pt_tourism_dataset[complete.cases(pt_tourism_dataset),]
```

### Renaming the key columns
```{r }
colnames(pt_tourism_dataset) <- c("month", "tourist")
print(pt_tourism_dataset)
```

### Convert to a time indexed table
```{r}

pt_tour_ts <- pt_tourism_dataset  %>%
  mutate(month = yearmonth(month)) %>% 
  as_tsibble(index = month)
print(pt_tour_ts)

```

### First Step: Visually assess the time series plot including the ACF and PACF to check if the process is stationary  
```{r message = FALSE, warning = FALSE}

pt_tour  <- pt_tour_ts  %>%
  filter(year(month) < 2020) %>%  
  mutate(tourist = tourist / 1000) %>%
  select(month, tourist)
gg_tsdisplay(pt_tour, tourist, plot_type = 'partial', lag = 36) +
  labs(title = "Portugal's Tourism Traffic (2008 - 2019)", y = "Number of Tourists (x1000)")
```

### Plot Results interpretation:

After plotting the graph and analyzing the plots, including the ACF and the PACF, there was an evidence of seasonality and a nonlinear trend. There is also an evidence of variance fluctuation, so we will stabilize the plot with a Box-Transformation and compare the plots with the stabilized variance and the original plot to know if the stabilization was necessary.

### Stabilizing the variance using a Box–Cox transform.
```{r}
library(dplyr)
library(feasts)
pt_tour %>% 
  gg_tsdisplay(log(tourist), plot_type = "partial", lag = 36) + labs(title = "Log Transformed", y = "")


```

### Plot Results interpretation:

The variance appeared to be stabilized but yet there is no significant change in the appearance of the plot when compared to the original plot. Therefore, we conclude that the change is negligible and we maintain the original data-set and proceed with the next step - Seasonal Difference. But firstly, we also have to test for the presence of a Unit Root. We will be installing a library urca which is used for the Augmented Dickey Fuller Test (ADF) to test for the trend and apply the appropriate difference.



## Apply unit root test with trend after applying box-cox trasformation:
```{r}
library(urca)
summary(ur.df(pt_tour$tourist, type = "trend", lags = 24))
```

### Interpreting the result for the ADF test 

We start from a high lag and descend to get the phi1 value of the last lag that falls below 10% significance. We started from lag 27 and settled at 24 lags which gave us a phi1 value of 0.000624 , which falls within the acceptable range of values. Using the GTS t-dig approach, the results of the ADF test showed the existence of a Unit Root, since '-1.811 ' is not more negative than '−3.43' (5% level), so we fail to reject the Null hypothesis. Thus we need at least one seasonal difference.

### Applying Seasonal differencing then re-checking stationarity.
```{r message = FALSE, warning = FALSE}
pt_tour %>%
  gg_tsdisplay(difference(tourist, 12), plot_type = "partial", lag = 36) +
  labs(title = "Seasonal Difference (12)", y = "")
``` 

### Seasonal difference results interpretation:

After taking the seasonal difference, the plot was analyzed and it still exhibited patterns of constitutionality and so, further tests will be needed to check for the presence of a unit root.

###  Unit root test with drift after applying: Seasonal Differencing
```{r}
pt_tour <- pt_tour %>%
  mutate(d12_tourist = difference(tourist, 12))
library(urca)
summary(ur.df(
  na.omit(pt_tour$d12_tourist),
  type = c('drift'),
  lags = 12
))
```

### Interpreting the result for the ADF test after seasonal differencing (lag = 12)

After analyzing the output of the ADF test with drift, the value of tau3 τ = −2.2048, which is above the 5% critical value −2.88. therefore we fail to reject the null hypothesis. for the next step we proceed to apply a first order difference.


### Applying Seasonal Differencing and First order Differencing

```{r message = FALSE, warning = FALSE}
pt_tour %>%
  gg_tsdisplay(difference(d12_tourist),
               plot_type = 'partial',
               lag = 36) +
  labs(title = "Seasonal + First order Differencing", y = "")
``` 

### First order difference report

After taking a further first order difference, the plot was analyzed and it exhibits a pattern of stationary. However, the decision is subject to a final ADF test to verify the absence of a unit root.

###  Unit root test after applying: Seasonal Differencing + First Order Differencing
```{r}
pt_tour <- pt_tour %>%
  mutate(dd_tourist = difference(d12_tourist))
library(urca)
summary(ur.df(
  na.omit(pt_tour$dd_tourist),
  type = c('none'),
  lags = 11
))
  
```
### Interpreting the result for the ADF test 

After analyzing the output of the ADF test, the ADF test statistic is τ =  −4.3116 which is less than −1.95,and now we can  reject the null hypothesis at the 5%. We proceed to model identification.

### Identification of the Mmdel through the assessment of the ACF and PACF

Based on the previous stationary plot, we will start with fitting several models.
We will then estimate the models, compare and assess the information criteria. We would start with using the PACF to select the non-seasonal part of the model and the ACF to select the seasonal part of the model. In our process, we removed the long-run trend and yearly seasonality through first and seasonal difference (d=1, D=1).

In the Seasonal and First Order difference plots, there is a spike in the PACF at lag 11 then cuts off, which give us the parameters for the non-seasonal part, suggesting an AR(1).
On the ACF plot, we can see a semblance of seasonality since there are spikes on lags 12, 24 and 36, which is an evidence of a requirement of more than 1 seasonal components, so we start with an MA(2) but can progress to an MA(3) if needed.

Consequently, this initial analysis suggests that a possible model for these data is an ARIMA(1,1,0)(0,1,2). We fit this model, along with some variations on it and compute the AICc values.

```{r}
pt_fit <- pt_tour %>% 
  model(
    sarima110_012 = ARIMA(tourist ~ pdq(1,1,0) + PDQ(0,1,2)),
    sarima111_012 = ARIMA(tourist ~ pdq(1,1,0) + PDQ(0,1,1)),
    sarima111_111 = ARIMA(tourist ~ pdq(1,1,1) + PDQ(1,1,1)),
    sarima110_112 = ARIMA(tourist ~ pdq(1,1,0) + PDQ(1,1,2)),
    sarima110_111 = ARIMA(tourist ~ pdq(1,1,0) + PDQ(1,1,1)),
    sarima111_112 = ARIMA(tourist ~ pdq(1,1,1) + PDQ(1,1,2))
  )


#checking the information criteria 

glance(pt_fit) %>% arrange(AICc) %>% select(.model:BIC)

```

### Information Criteria Report:

After estimating the candidate SARIMA specifications from the AC F/PACF, we compared their information criteria. Consistent with the Box–Jenkins procedure, we use AICc as the primary selection metric and BIC as a parsimony check among models with adequate residuals. The model SARIMA(1,1,0)(1,1,2)[12] achieves the lowest AICc (1376.17) and lowest BIC (1390.07) with the highest log-likelihood (–682.85), indicating the best balance of fit and complexity among the candidates. We therefore select it as the preferred model for forecasting.

### Diagnostic check plot for residuals

From the analysis table of the estimated models, we select the model SARIMA(1,1,0)(1,1,2) according to the information criteria. Then we proceed to analyze the residuals.

```{r message = FALSE, warning = FALSE}
pt_fit %>%
  select(sarima110_112) %>%
  gg_tsresiduals()
```

### Diagnostics Check Report

After analyzing the values of the ACF plot of the residuals, there was no significant pattern left to model; as they look random. The histogram of the residual plots also approximates a normal distribution. Following the process, there is a need to test for the presence of autocorrelation between the residuals through Ljung-Box and Box-Pierce tests.

### Statistical Tests of the Residuals
```{r}
#Ljung-Box test:
augment(pt_fit) %>%
filter(.model=='sarima110_112') %>%
features(.innov, ljung_box, lag = 36)
```

```{r}
#Box-Pierce test:
augment(pt_fit) %>%
filter(.model=='sarima110_112') %>%
features(.innov, box_pierce, lag = 36)
```

## Report from the Statistical Analysis test of the Residuals

Following the output of the Ljung-Box and Box-Pierce residuals test, it is evident that the p-values for the tests are all greater than 5%, so we don’t reject the null hypothesis. This indicates that there is no statistical evidence of the presence of autocorrelation between the residuals. 

Before finalizing our forecast, we will further explore whether an alternative forecasting method might yield a better statistical fit for the data. To do this, we will specifically examine the Exponential Smoothing State Space (ETS) model.

```{r }
# Fitting ETS model on the same series

ets_model <- pt_tour %>%
  model(ETS = ETS(tourist))

# Extract info criteria for both SARIMA(1,1,0)(1,1,2) and ETS
info_compare <- bind_rows(
  glance(pt_fit %>% select(sarima110_112)),
  glance(ets_model)
) %>%
  select(.model, sigma2, log_lik, AIC, AICc, BIC)

info_compare
```
### Comparison with Alternative Forecasting Method (ETS)

The 'sarima110_112' model achieved substantially lower information criteria than the ETS alternative, confirming its superior fit.

This indicates that SARIMA better captures the stochastic seasonal structure and autocorrelation dynamics of the tourism data, whereas ETS models seasonality through smoother deterministic components.
Therefore, SARIMA remains the preferred model for forecasting Portugal’s tourism series.

## Tourism Series Forecast

### Out-of-Sample Forecast Evaluation (2019 Test Set)
```{r}
forecast(pt_fit, h=36) %>%
filter(.model=='sarima110_112') %>%
autoplot(pt_tour) +
labs(title = "Portugal's Tourism Traffic",
y="number of tourists (x1000)")

```

### Validating the accuracy of the selected model

After forecasting the series based on the model, it is imperative to validate the model's accuracy. We  need to test how well each model can predict data it has not seen. To do this, we split the series into a training set (all observations up to 2018) and a test set (2019 only). Each candidate model is estimated using only the training data, so it “learns” patterns such as trend, seasonality, and short-term dependence without looking at 2019. We then generate forecasts for 2019 and compare those forecasts to the actual observed values in 2019. This is called out-of-sample evaluation, and it’s more honest than just checking in-sample fit, because it tests true predictive performance. After producing the forecasts, we can both visually inspect how close the forecast lines are to the real 2019 values, and compute accuracy metrics (ME, RMSE, MAE, MPE, MAPE, MASE, RMSSE, ACF1) to quantify forecast error. The model with the lowest forecast error on the 2019 test set is the model we’ll consider most reliable for future forecasting.

### In-Sample Forecast and Diagnostic Evaluation
```{r}
#Training data
pt_train <- pt_tour %>%
  filter(year(month) < 2019)
#Test data
pt_test <- pt_tour %>%
  filter(year(month) >= 2019, year(month) <= 2019)
#Fit models in training data
pt_fit <- pt_train %>%
  model(
    sarima111_112 = ARIMA(tourist ~ pdq(1,1,1) + PDQ(1,1,2)),
    sarima110_112 = ARIMA(tourist ~ pdq(1,1,0) + PDQ(1,1,2)),
    sarima110_011 = ARIMA(tourist ~ pdq(1,1,0) + PDQ(0,1,1)),
  )
#Forecast in test data
pt_fc <- pt_fit %>%
  forecast(pt_test)
#Plot results
pt_fc %>%
  autoplot(bind_rows(pt_train, pt_test), level = NULL) +
  labs(y = '', title = 'Tourism') +
  guides(colour = guide_legend(title = "Forecast"))
```

```{r }
#Comparing the model's accuracy
fabletools::accuracy(pt_fc, dplyr::bind_rows(pt_train, pt_test))

```

### Accuracy Report and Conclusion

To justify the model selection, the SARIMA(1,1,0)(1,1,2)[12] model was selected as the final forecasting model based on its strong out-of-sample performance and adherence to the Box–Jenkins principles.
It yielded the lowest overall forecast errors (RMSE = 50.05, MAE = 41.22, MAPE = 3.06%) and produced residuals that are approximately white noise, with no significant autocorrelation (ACF₁ = 0.31).
Following the Box–Jenkins criterion of selecting the most parsimonious and statistically adequate model, SARIMA(1,1,0)(1,1,2)[12] provides the best balance between model simplicity, goodness of fit, and forecasting accuracy for Portugal’s tourism demand series.

***Future improvements*** could involve incorporating external regressors such as international flight arrivals or macroeconomic indicators.



